Convergence rate of MAP estimates \\ for the exponential family

### 2021-09-22 main.tex
Precise variance n=1 case

As for the MAP, we did not manage to get an asymptotically tight upper bound. Nevertheless, using inequality~\eqref{eq:log_bound}, we did get an interesting upper bound. 
To start, let us recall that the MAP of $\cN(0,\mu_*)$ is $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i X_i^2}{n_0 + n}$.
Its expectation is simply $\mu_n= \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$.
We now present a lemma on its expected inverse.
\begin{lemma}[Expected MAP Natural Parameter]
	Let's define $a=n_0 \frac {\mu_0 }{ \mu^* }$. For any $n\geq 2$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
	\begin{align}
		\frac{\mu^*}{\mu_n} = \frac{n_0+n}{a+n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\natp^*}} 
		\leq \frac{n_0+n}{a+n-2} \; .
	\label{eq:theta_expectation_bound}
	\end{align}
\end{lemma}
Note that the lower bound is a simple consequence of the convexity of the inverse function. 
Using this lemma, along with the log upper bound on the logarithm~\eqref{eq:log_bound}, we get a convergence rate for the MAP. 
\begin{example}
\begin{theorem}[MAP Bound]
 Defining $a=n_0 \frac {\mu_0 }{ \mu^* }$, the expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
 \begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
	\leq \begin{cases}
		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
		\bregmanconj( \mu_*; \mu_0) + \inv{a} \ \text{if}\ n=1,\\
		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{\frac{a+n-2}{n_0}(1 + \frac{n}{n_0} )} \ \text{if}\  n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
	\end{cases}
	\label{eq:MAP_rate}
\end{equation}
\end{theorem}
\end{example}
Using the same bound on the log~\eqref{eq:log_bound}, we  get a loose bound on the MLE, which is useful for comparison purposes.
We relate these loose convergence rates of MAP and MLE in Figure~\ref{fig:MAP_rate}.
\begin{example}
\begin{corollary}[MLE Loose Bound]
\begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
	\leq \inv{n-2} \; .
	\label{eq:MLE_loose_rate}
\end{equation}
\end{corollary}
\end{example}

@@

@@

\subsection{Primal Expectation Pivot}
Let $\tilde \theta_n := \expect{\hat \theta_n}$ and $\tilde \mu_n = \nabla \logpart(\tilde \theta_n )$ be the primal and dual parameters of the primal expectation of the MAP.
As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like
\begin{align}
	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
	&= \bias{\bregmanconj(\mu^* ; \tilde \mu_n)}
	+ \variance{\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}} \\
	\expect{\bregman(\MAPt; \natp^*)} 
	&= \bias{\bregman(\E[\MAPt]; \natp^*)}
	+ \variance{\expect{\bregman(\MAPt; \E[\MAPt])}}  \; .
\end{align}

@@

\begin{lemma}
Taking  $\MAPt$ and $\MAPm$ as the primal and dual parameters of the MAP estimate, and $\tilde \mu_n$ the primal expectation and $\mu_n$ the dual expectation, the following identities hold
	\begin{align}
	\Cov(\hat \natp_n ; \MAPm )  
	&=\expect{\bregmanconj(\MAPm ; \mu_n)
	+ \bregmanconj(\mu_n ; \MAPm)} \\
	&=\expect{\bregmanconj(\MAPm ; \tilde \mu_n)
	+ \bregmanconj(\tilde \mu_n ; \MAPm)} \; .
\end{align}
\end{lemma}
\begin{proof}
By the three point identity (or plain derivation), we have 
\begin{align}
	\bregmanconj(\MAPm ; \mu_n)
	+ \bregmanconj(\mu_n ; \MAPm)
	= \langle \nabla \conj (\hat \mu_n) - \nabla \conj(\mu_n) ;  \hat \mu_n - \mu_n \rangle 
	= \langle \MAPt - \theta_n;  \MAPm- \mu_n \rangle 
\end{align}
Taking the expectation,  the constant $\theta_n$ disappears since $\expect{\hat \mu_n - \mu_n}=0$.
\begin{align}
	\expect{\langle \MAPt - \theta_n;  \MAPm- \mu_n \rangle } 
	 = \expect{\langle \MAPt; \MAPm \rangle} - \langle \expect{\MAPt} ; \MAPm \rangle
	= \Cov(\hat \natp_n ; \hat \mu_n ) 
\end{align}
which is the first identity. We get the second identity by symmetry between the primal and the dual.
\end{proof}

@@

\paragraph{Expansion}
We can expand  $\hat \mu_n - \mu_n$ to get another formulation of this covariance
\begin{align}
	\expect{\langle \hat \natp_n ; \hat \mu_n - \mu_n \rangle}
	&= \expect{\langle \hat \natp_n ; \frac{\sum_i T(X_i) - \mu^*}{n+n_0} \rangle}  \\
	& = \frac{n}{n+n_0} \expect{\langle \hat \natp_n ; T(X_0) - \mu^*\rangle}
	= \frac{n}{n+n_0} \Cov(\hat \natp_n ; T(X_0) ) 
\end{align}
where we used the property that all data points play a symmetric role to remove the sum over $i$.

@@

\paragraph{Mirror Descent.}
Let us recall mirror descent iteration
\begin{align}
	\natp_{t+1} 
	&:= \argmin_\natp \lr \linear_f(\natp; \natp_t) + \bregman(\natp ; \natp_t)  \\
	& = \nabla \conj (\nabla \logpart(\natp_t) - \lr \nabla f(\natp_t))
\end{align}
where $\linear_f(\natp; \natp_t) = f(\natp_t) + \langle \nabla f(\natp_t), \natp - \natp_t \rangle$ is the linear approximation of $f$ in $\natp_t$ evaluated at $\natp$. Solving this problem require solving problems of the form $\argmin_\natp - \langle c, \natp \rangle + \logpart(\natp)$, eg computing the convex conjugate of $\logpart$. 
Note that finding $\natp_*$ is done with 1 step of mirror descent. Indeed plugging in definitions of $f$ and $\mu$ yields
\begin{align}
	&\mu_{t+1}  = \mu_t - \lr (\mu_t - \mu_*) \\
	\implies &\mu_t  = \mu_* + (1- \lr)^t (\mu_0 - \mu_*)
\end{align}
which shows exponential convergence and 1-step convergence when $\lr =1$.

@@

Stochastic Mirror Descent (SMD) was studied under relative smoothness assumptions in 2 papers : \citet{hanzely2018fastest} and \citet{dragomir2021fast}.

@@

\citet{hanzely2018fastest} got convergence rates in the smooth strongly-convex case with tail averaging. With constant step-size, they proved linear convergence down to a variance ball. With step-size $\gamma_t = n_0 + t$, they proved a rate $\tilde O(\inv{t})$. These results match the rates for standard SGD.
Let us have a look at their variance assumption \citep[Assumption 5.1]{hanzely2018fastest}.
Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate. Then the variance bound $\sigma^2$ is an upper bound on the covariance between the gradient update  $-g_t$ and the descent direction $\natp_{t+1} - \natp_t = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t ) - \natp_t $ that should hold for all time steps  
\begin{align}
	\lr_t \sigma^2 \geq 
	&\Cov(- g_t,\natp_{t+1} - \natp_t | X_{1 \dots t})  \label{eq:hanzely_hypothesis} \\
	&= \expect{\langle - g_t, \natp_{t+1} \rangle}  - \langle \E[-g_t], \E[\natp_{t+1}] \rangle   \\
	 &= \expect{\langle \nabla f(\natp_t) - g_t , \natp_{t+1}  \rangle}\\
	 &= \expect[X_t]{\bregman(\natp_{t+1} ; \bar \natp_{t+1} ) + \bregman(\bar \natp_{t+1} ; \natp_{t+1} ) }
\end{align}
where $\gamma_t$ is the step-size, expectations are conditional on the past, and $\bar \natp_{t+1}$ is the theoretical output of a deterministic Bregman gradient step. Remark that when we plug in $\conj = \|.\|^2$, we recover the gradient variance typical of SGD.

In our setting, 
\begin{align}
	\nabla f(\natp_t) - g_t =  T(X_{t+1})	- \E[T(X)]
\end{align}
so that $\sigma$ is really a bound on the covariance of the sufficient statistics with another variable
\begin{align}
	&\Cov( T(X_{t+1}), \natp_{t+1}| X_{1\dots t})
	\leq \lr_t\sigma^2 \\
	=& \Cov(T(X_{t+1}), \nabla\conj \left(\lr_t T(X_{t+1}) + (1-\lr_t ) \mu_t \right) | X_{1\dots t}) \;.
\end{align}
In other words, we need for all $\lr\in(0,\inv{n})$, and a certain subset of $\mu$,
\begin{align}
\label{eq:hypvariance}
	\boxed{
	\Cov_X \bigg(T(X), \nabla \conj (\lr T(X) + (1-\lr) \mu)\bigg	) 
	\leq \lr \sigma^2 \; .
	}
\end{align}

\begin{example}
Unfortunately this bound is not trivially satisfied, even for simple exponential family members, such as centered gaussians $\cN(0,\sigma^2)$, when we are aiming to estimate the variance.
The covariance  \eqref{eq:hypvariance} we want to bound is equal to
\begin{align}
	\half \expect{\frac{\sigma_*^2 - X^2 }{ X^2 \lr + (1-\lr) \sigma_0^2 }}
\end{align}
When $\sigma_0 \rightarrow 0$, as might happen, this value explodes.
\end{example}

@@

\subsection{Transferring SMD's covariance assumption to MAP estimates}

\paragraph{MLE Analysis.}
When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, we get
\begin{multline}
	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ;\hat \mu) }
	=\conj(\mu_*) \overbrace{- \E[\conj(\hat \mu)]}^{\leq - \conj(\E[\hat \mu])} \\
	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance}
	 	\; \text{with}\; \mu=\inv{n-1}\smallsum_{j\neq i} T(x_j)  } ]  \leq \frac{\sigma^2}{n}
\label{eq:MLE_bound}
\end{multline}
where we used the decomposition $\hat \mu = \inv{n}T(X_i) + (1-\inv{n}) \inv{n-1} \sum_{j\neq i}T(X_j)$ to apply~\eqref{eq:hypvariance}. 
In words, this variance assumption on $T(X)$ and $\conj$ immediately gives us a bound on the suboptimality.

This is too simple to be true, as shown on our trailing example.
In particular we show that it cannot hold for all possible values of $\mu$ within the double expectation in~\eqref{eq:MLE_bound}.
The mean parameter $\mu$ take values arbitrarily close from $0$, which makes the covariance explode.
\RLP{should we get a functional bound $\sigma^2(\mu)$ instead ? or split between $\mu$ smaller or larger than $\varepsilon$.}

\paragraph{MAP Analysis.}
We can also apply this result to the MAP estimate $\hat \mu = \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n}$ but I did not manage to reach a satisfying conclusion about the bias. 
Note $\tilde \mu = \E[\hat \mu] = \frac{n_0 \mu_0 + n \mu_*}{n_0+n}$ and $\gamma_0= \frac{n_0}{n_0+n}$ and the step-size $\gamma = \inv{n_0 + n}$ .
\begin{align}
	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ;\hat \mu) } 
	&\leq \conj(\mu_*) - \conj(\tilde \mu) + \gamma_0 \langle \mu_0 - \mu_* ; \E[\hat \natp] \rangle + (1 - \gamma_0) \gamma \sigma^2  \\
	& = \bregmanconj (\mu_* ; \tilde \mu) - \langle \tilde \mu - \mu_* ;    \tilde \natp - \E[\hat \natp] \rangle + \frac{n}{(n_0+n)^2}\sigma^2
\end{align}
so we recover the $O(n^{-1})$ for the variance and we are still looking for $O(n^{-2})$ rate for the bias. The Bregman term in the bias should be in this spirit, given a quadratic approximation
\begin{align}
	 \bregmanconj (\mu_* ; \tilde \mu)  
	 = \bregmanconj (\mu_* ; \mu_* + \gamma_0 (\mu_0 - \mu_*) ) \approx \half \gamma_0^2 \norm{\mu_0 - \mu_*}^2_{\mSigma_*^{-1}}
\end{align}
where $\mSigma_* = \Cov_{\mu_*}(T(X))$. 
The order of terms $\bregmanconj (\mu_* ; \tilde \mu)  \neq \bregmanconj (\tilde \mu ; \mu_* ) $ only affects third order terms.
That's for the bregman, but the scalar product is harder to bound. Hopefully it could be positive. If not we know that $\tilde \mu - \mu_* = \frac{n_0}{n_0+n} (\mu_0 - \mu_*) = O(n^{-1})$ , and we can hope that the same holds for $\tilde \natp - \E[\hat \natp] = \nabla \conj(\E[\hat \mu]) - \E[\nabla\conj (\hat \mu)]$, which kinda measures the non-linearity, or the curvature of $\conj$. Using a simple cauchy-schwartz, and the hessian of the conjugate, we might be able to get something.

\paragraph{Relative Continuity}
\citet{lu2019relative} define relative continuity to tackle non-differentiable functions -- eg Hinge loss. These functions are typically assumed to be Lipschitz, but that is not always true -- eg squared Hinge loss. Relative continuity generalizes Lipschitz-ness by upper bounding the gradient norm (or expected norm of stochastic gradients) with the ratio between a Bregman and $\ell^2$ distance. Although this hypothesis is far from intuitive, it does yield convergence rates very similar to typical SGD analysis. In this work we are first concerned with strictly convex differentiable functions $\logpart(\natp)$, to ensure a clean bijection between natural and mean parameters.

\paragraph{Bound on the trajectory.}
Bound \eqref{eq:hypvariance} might not hold in general, but it can along the trajectory. Can we prove it using an inductive method such as \citet[Appendix A]{lacoste2012simpler}?
No we cannot because $\mu$ takes every possible values as a random variable within the expectation.

@@

@@

\citet{kunstner2020homeomorphic} show with an exquisite elegance that Expectation-Maximization in exponential families can be cast as mirror descent on the negative log-likelihood. Then they use recent mirror descent rates based on the idea of relative smoothness \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively} to show convergence of EM in KL-divergence, including for the ubiquitous Gaussian Mixture model, which had resisted bounding attempts for 50 years.

@@

@@

\subsection{Proof of Expected MAP Natural Parameter}
\begin{lemma}[Expected MAP natural parameter]
	The expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
	\begin{align}
		\frac{\mu^*}{\mu_n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\natp^*}} 
		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n+n_0}} \; .
	\end{align}
\end{lemma}

>>>>

As for the MAP, we did not manage to get an asymptotically tight upper bound. Nevertheless, using inequality~\eqref{eq:log_bound}, we did get an interesting upper bound. 
To start, let us recall that the MAP of $\cN(0,\mu_*)$ is $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i X_i^2}{n_0 + n}$.
Its expectation is simply $\mu_n= \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$.
We now present a lemma on its expected inverse.
\begin{lemma}[Expected MAP Natural Parameter]
	Let's define $a=n_0 \frac {\mu_0 }{ \mu^* }$. For any $n\geq 1$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
	\begin{equation}
		\frac{\mu^*}{\mu_n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\natp^*}} 
		\leq \frac{n_0 +n}{a+ (n-2)_+}
		\label{eq:theta_expectation_bound}
	\end{equation}
	where $(x)_+ = \max(0,x)$.
\end{lemma}
Note that the lower bound is a simple consequence of the convexity of the inverse function. 
Using this lemma, along with the log upper bound on the logarithm~\eqref{eq:log_bound}, we get a convergence rate for the MAP. 
\begin{example}
\begin{theorem}[MAP Bound]
 Let us  define
 \begin{align}
	b = \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; .
 \end{align}
The expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
 \begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b\ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\mu_0}{\mu^*} +n-2} + b \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}
\end{equation}
\end{theorem}
\end{example}
This inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\mu_0}{\mu^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.  For instance, when $n_0=1$, it encourages us to set $\mu_0 = 2 \mu^*$.
Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$. We are basically losing a factor 2 compared to $1/2n$.

Using the same bound on the log~\eqref{eq:log_bound}, we  get a loose bound on the MLE, which is useful for comparison purposes.
We relate these loose convergence rates of MAP and MLE in Figure~\ref{fig:MAP_rate}.
\begin{example}
\begin{corollary}[MLE Loose Bound] When $n\geq 2$, the MLE expected loss is upper bounded as
\begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
	\leq \inv{n-2} \; .
	\label{eq:MLE_loose_rate}
\end{equation}
\end{corollary}
\end{example}

@@

\TODO{Tonight ! add equations for the full gaussian. I can dig them ou of my previous notebook.}

\TODO{add equations for other families, eg gaussian with mean and variance, or gamma or Dirichlet, or Categorical..}

@@

\subsection{Primal Expectation Pivot}
Let $\tilde \theta_n := \expect{\hat \theta_n}$ and $\tilde \mu_n = \nabla \logpart(\tilde \theta_n )$ be the primal and dual parameters of the primal expectation of the MAP.
As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like
\begin{align}
	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
	&= \bias{\bregmanconj(\mu^* ; \tilde \mu_n)}
	+ \variance{\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}} \\
	\expect{\bregman(\MAPt; \natp^*)} 
	&= \bias{\bregman(\E[\MAPt]; \natp^*)}
	+ \variance{\expect{\bregman(\MAPt; \E[\MAPt])}}  \; .
	\label{eq:primal_pivot}
\end{align}

@@

\begin{lemma}
\label{lem:covariance}
Taking  $\MAPt$ and $\MAPm$ as the primal and dual parameters of the MAP estimate, and $\tilde \mu_n$ the primal expectation and $\mu_n$ the dual expectation, the following identities hold
	\begin{align}
	\Cov(\hat \natp_n ; \MAPm )  
	&=\expect{\bregmanconj(\MAPm ; \mu_n)
	+ \bregmanconj(\mu_n ; \MAPm)} \\
	&=\expect{\bregmanconj(\MAPm ; \tilde \mu_n)
	+ \bregmanconj(\tilde \mu_n ; \MAPm)} \; .
\end{align}
\end{lemma}
\begin{proof}
By the three point identity (or plain derivation), we have 
\begin{align}
	\bregmanconj(\MAPm ; \mu_n)
	+ \bregmanconj(\mu_n ; \MAPm)
	= \langle \nabla \conj (\hat \mu_n) - \nabla \conj(\mu_n) ;  \hat \mu_n - \mu_n \rangle 
	= \langle \MAPt - \theta_n;  \MAPm- \mu_n \rangle 
\end{align}
Taking the expectation,  the constant $\theta_n$ disappears since $\expect{\hat \mu_n - \mu_n}=0$.
\begin{align}
	\expect{\langle \MAPt - \theta_n;  \MAPm- \mu_n \rangle } 
	 = \expect{\langle \MAPt; \MAPm \rangle} - \langle \expect{\MAPt} ; \MAPm \rangle
	= \Cov(\hat \natp_n ; \hat \mu_n ) 
\end{align}
which is the first identity. We get the second identity by symmetry between the primal and the dual.
\end{proof}

@@

\paragraph{Expansion}
We can expand  $\hat \mu_n - \mu_n$ to get another formulation of this covariance
\begin{align}
	\expect{\langle \hat \natp_n ; \hat \mu_n - \mu_n \rangle}
	&= \expect{\langle \hat \natp_n ; \frac{\sum_i T(X_i) - \mu^*}{n+n_0} \rangle}  \\
	& = \frac{n}{n+n_0} \expect{\langle \hat \natp_n ; T(X_0) - \mu^*\rangle} \\
	&= \frac{n}{n+n_0} \Cov(\hat \natp_n ; T(X_0) ) 
	\label{eq:covariance_expansion}
\end{align}
where we used the property that all data points play a symmetric role to remove the sum over $i$.

@@

\paragraph{Mirror Descent.}
Let us recall mirror descent iteration
\begin{align}
	\natp_{t+1} 
	&:= \argmin_\natp \lr \linear_f(\natp; \natp_t) + \bregman(\natp ; \natp_t)  \\
	& = \nabla \conj (\nabla \logpart(\natp_t) - \lr \nabla f(\natp_t))
\end{align}
where $\linear_f(\natp; \natp_t) = f(\natp_t) + \langle \nabla f(\natp_t), \natp - \natp_t \rangle$ is the linear approximation of $f$ in $\natp_t$ evaluated at $\natp$. Solving this problem require solving problems of the form $\argmin_\natp - \langle c, \natp \rangle + \logpart(\natp)$, eg computing the convex conjugate of $\logpart$. 
Note that finding $\natp_*$ is done with 1 step of mirror descent with step-size 1.
Indeed plugging in definitions of $f$ and $\mu$, and assuming constant step-size $\lr$ yields
\begin{align}
	&\mu_{t+1}  = \mu_t - \lr (\mu_t - \mu_*) \\
	\implies &\mu_t  = \mu_* + (1- \lr)^t (\mu_0 - \mu_*)
\end{align}
which shows exponential convergence and 1-step convergence when $\lr =1$.

@@

We know of 2 papers that study Stochastic Mirror Descent (SMD) under relative smoothness assumptions -- \citet{hanzely2018fastest} and \citet{dragomir2021fast}. We are now going to review these works.

@@

\paragraph{General Approach.} 
\citet{hanzely2018fastest} got convergence rates in the smooth strongly-convex case with tail averaging. With constant step-size, they proved linear convergence down to a variance ball. With step-size $\gamma_t = n_0 + t$, they proved a rate $\tilde O(\inv{t})$. These results match the rates for standard SGD.
Let us have a look at their variance assumption \citep[Assumption 5.1]{hanzely2018fastest}.
Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate. Then the variance bound $\sigma^2$ is an upper bound on the covariance between the gradient update  $-g_t$ and the next iterate $\natp_{t+1} = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t )$ that should hold for all time steps  
\begin{align}
	\lr_t \sigma^2 \geq 
	&\Cov(- g_t,\natp_{t+1} | X_{1 \dots t})  \label{eq:hanzely_hypothesis} \\
	 &= \expect{\langle \nabla f(\natp_t) - g_t , \natp_{t+1}  \rangle}
\end{align}
where $\gamma_t$ is the step-size, and expectations are conditional on all of the past. 
Remark that when we plug in $\conj = \|.\|^2$, we recover the gradient variance typical of SGD. 
If we introduce $\bar \natp_{t+1}=\nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t  \nabla f(\natp_t) )$ the theoretical output of a deterministic Bregman gradient step, then following Lemma~\ref{lem:covariance} we can rewrite this covariance as the expectation of symmetrized Bregman between stochastic and deterministic iterates $\expect[X_t]{\bregman(\natp_{t+1} ; \bar \natp_{t+1} ) + \bregman(\bar \natp_{t+1} ; \natp_{t+1} ) }$.
This view is useful to understand that we are measuring the variance of each iteration, which is necessarily positive.

\paragraph{MAP Special Case.}
In our setting, 
\begin{align}
	\nabla f(\natp_t) - g_t =  T(X_{t+1})	- \E[T(X)]
\end{align}
so that $\sigma$ is really a bound on the covariance of the sufficient statistics $T(X_{t+1})$ with the next iterate $\natp_{t+1}$, which itself depends on $T(X_{t+1})$ via a coefficient $\lr_t$ and a non-linearity $\nabla\conj$.
\begin{align}
	&\Cov( T(X_{t+1}), \natp_{t+1} | X_{1\dots t})
	\leq \lr_t\sigma^2 
	\label{eq:covariance_assumption} \\
	=& \Cov(T(X_{t+1}), \nabla\conj \left(\lr_t T(X_{t+1}) + (1-\lr_t ) \mu_t \right) | X_{1\dots t}) \;.
\end{align}
In other words, we need for all $\lr_t \in [0; \lr_0]$, and the subset of $\mu$ that can be reached by on the trajectory $\mu_t$,
\begin{align}
\label{eq:hypvariance}
	\boxed{
	\Cov_X \bigg(T(X), \nabla \conj (\lr T(X) + (1-\lr) \mu)\bigg	) 
	\leq \lr \sigma^2 \; .
	}
\end{align}

\begin{example}
Unfortunately this bound is not trivially satisfied, even for simple exponential family members.
For instance, centered gaussians variance estimation $\cN(0,\mu^*)$ does not verify \eqref{eq:hypvariance}.
Indeed the covariance  \eqref{eq:hypvariance} we want to bound is equal to
\begin{align}
	\half \expect[X\sim \cN(0,\mu^*)]{\frac{\mu*^2 - X^2 }{ X^2 \lr + (1-\lr) \mu^2 }}
	= \inv{\lr} (1+a) \expect[X \sim \cN(0,1)]{\inv{a + X^2}} - \inv{\lr} 
\end{align}
where $a = \frac{1-\lr}{\lr} \frac{\mu}{\mu^*}$
When $\sigma_0 \rightarrow 0$, as might happen, this value explodes.
\end{example}

\paragraph{Straightforward analysis}
When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, we get
\begin{multline}
	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ;\hat \mu) }
	=\conj(\mu_*) \overbrace{- \E[\conj(\hat \mu)]}^{\leq - \conj(\E[\hat \mu])} \\
	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance}
	 	\; \text{with}\; \mu=\inv{n-1}\smallsum_{j\neq i} T(x_j)  } ]  \leq \frac{\sigma^2}{n}
\label{eq:MLE_bound}
\end{multline}
where we used the decomposition $\hat \mu = \inv{n}T(X_i) + (1-\inv{n}) \inv{n-1} \sum_{j\neq i}T(X_j)$ to apply~\eqref{eq:hypvariance}. 
In words, this variance assumption on $T(X)$ and $\conj$ immediately gives us a bound on the suboptimality.

This is too simple to be true, as shown on our trailing example.
In particular we show that it cannot hold for all possible values of $\mu$ within the double expectation in~\eqref{eq:MLE_bound}.
The mean parameter $\mu$ takes values arbitrarily close from $0$, which makes the covariance explode.

\paragraph{Bias-Variance Perspective.}
As shown in equation~\eqref{eq:covariance_expansion}, covariance~\eqref{eq:covariance_assumption} is an upper bound on the variance term in either the primal \eqref{eq:primal_pivot} or dual \eqref{eq:bias-mixed-variance} expectation pivot decompositions
\begin{align}
	\Cov(\MAPm, \MAPt)  = \frac{n}{n+n_0} \expect[X_{1:n-1}]{\underbrace{\Cov(\MAPt ; T(X_{n}) | X_{1:n-1} )}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance}}}
	\leq  \frac{n \sigma^2}{(n+n_0)^2}
\end{align}
so we recover a $O(n^{-1})$ upper bound on the variance term and we are still looking for the $O(n^{-2})$ rate for the bias term. It is simpler to consider the primal pivot, but still hard to say anything about $\bregmanconj(\mu^* ; \tilde \mu_n) = \bregman(\E[\MAPt]; \natp^*)$ without further assumptions.

@@

@@

\subsection{Discussion}

\paragraph{Bound on the trajectory.}
Bound \eqref{eq:hypvariance} might not hold in general, but it can along the trajectory. Can we prove it using an inductive method such as \citet[Appendix A]{lacoste2012simpler}? In the case of the Gaussian variance, we cannot because $\mu$ takes every possible values as a random variable within the expectation.

@@

\citet{kunstner2020homeomorphic} show with an exquisite elegance that Expectation-Maximization in exponential families can be cast as mirror descent on the negative log-likelihood. Then they use recent mirror descent rates based on the idea of \textit{relative smoothness} \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively} to show convergence of EM in KL-divergence, including for the ubiquitous Gaussian Mixture model, which had resisted bounding attempts for 50 years.

Relative smoothness gave rise to sibling hypothesis, such as \textit{relative continuity} \citep{lu2019relative},  which was used to characterize non-differentiable functions -- eg Hinge loss. These functions are typically assumed to be Lipschitz, but that is not always true -- eg squared Hinge loss. Relative continuity generalizes Lipschitz-ness by upper bounding the gradient norm (or expected norm of stochastic gradients) with the ratio between a Bregman and the $\ell^2$ distance. Although this hypothesis is far from intuitive, \citet{lu2019relative} obtained convergence rates for mirror descent that are very similar to typical SGD analysis. In this work we are concerned with strictly convex differentiable functions $\logpart(\natp)$, to ensure a clean bijection between natural and mean parameters. Consequently, we do not further investigate relative continuity.

@@

\section{Other Exponential Families}
\label{app:expfamily}

\subsection{Full Gaussian $\cN(m, \sigma^2)$}
Take the sufficient statistic $T(X) = (X, X^2)$. Then the mean parameter is $\mu = \E[T(X)] = ( m , m^2 + \sigma^2) = (\mu_1, \mu_2) $.
The log-density is written
\begin{align}
	\log p(X) 
	&= -\frac{(X-m)^2}{2 \sigma^2} - \half \log \sigma^2 + \cst \\
	&=  \frac{m}{\sigma^2} X -\inv{2 \sigma^2} X^2 - \frac{m^2}{2 \sigma^2} - \half \log \sigma^2 \\
	&=  \natp_1 X + \natp_2 X^2  - \logpart(\natp) \; .
\end{align}
From this formula, we can identify the duality map
\begin{align}
	\natp_1 &= \frac{m}{\sigma^2} = \frac{\mu_1}{\mu_2 - \mu_1^2} \\
	\natp_2 &= -\inv{2 \sigma^2} =  \frac{ - 1}{2(\mu_2 - \mu_1^2)}  < 0
\end{align}
and we can also explicit the log-partition as a function of the natural parameters
\begin{align}
	\logpart(m, \sigma) &= \frac{m^2}{2 \sigma^2} + \half \log \sigma^2 \\
	\logpart(\natp) &= \frac{\natp_1^2}{-4\natp_2} - \half \log(-\natp_2) \; .
\end{align}
We can perform a sanity check by deriving $\logpart$ and verifying that its gradient gives $\mu(\natp)$
\begin{align}
	\mu_1 &= \frac{\natp_1}{-2\natp_2}  \\
	\mu_2 & = \left( \frac{\natp_1}{-2\natp_2} \right)^2 + \inv{-2 \natp_2} \; .
\end{align}
Now let us discover the entropy
\begin{align}
	\conj (\mu) &= \mu_1 \natp_1 + \mu_2 \natp_2  - \logpart(\natp(\mu)) \\
	& = \frac{\mu_1^2}{\mu_2 - \mu_1^2}  - \frac{\mu_2}{2(\mu_2 - \mu_1^2)} - \frac{\mu_1^2}{2(\mu_2 - \mu_1^2)}  - \half \log(\mu_2 - \mu_1^2) \\
	&= \half \frac{\mu_1^2 - \mu_2}{\mu_2 - \mu_1^2} - \half \log(\mu_2 - \mu_1^2) \\
	& = - \half \log(\mu_2 - \mu_1^2) + \cst \; .
\end{align}
In this case, the Bregman divergences induced by the log-partition or the entropy are ugly, so we will not focus on them.

@@

\subsection{Proof of Expected MAP Natural Parameter}
Let us introduce a variable characterizing the importance of the prior
\begin{important}
	\begin{align}
			a=n_0 \frac{\mu_0}{\mu^*} \; .
	\end{align}
\end{important}
\begin{lemma}[Expected MAP natural parameter]
	The expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
\begin{equation}
		\frac{\mu^*}{\mu_n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\natp^*}} 
		\leq \begin{cases}
		\frac{n_0 +n}{a+n-2} \text{ when } n\geq 2,\\
		\frac{n_0 +1}{a} \text{ when } n=1.
		\end{cases}
		\label{eq:lemma_natp_bound}
	\end{equation}
\end{lemma}
